---
title: "Response Curve Theory and Data Requirements"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Response Curve Theory and Data Requirements}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4,
  warning = FALSE,
  message = FALSE
)
```

```{r libs}
library(mrmopt)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
```

## The Saturation Problem

Media channels do not deliver returns in proportion to spend. The first
dollar spent on a channel is typically the most efficient — it reaches
the most responsive audience, captures the highest-intent search
queries, or occupies the cheapest inventory. As spend increases, each
additional dollar reaches progressively less responsive audiences,
drives incrementally smaller lifts, and encounters diminishing returns.

This behaviour is well-described by an **S-shaped saturation curve**:
response rises steeply at first, passes through an inflection point, and
gradually levels off toward an asymptotic ceiling. The shape encodes two
important economic realities:

1.  **Diminishing marginal returns** — the incremental KPI generated by
    an additional dollar of spend falls as total spend grows.
2.  **A capacity or addressable market ceiling** — there are only so
    many consumers in the market, only so much demand to capture.

Linear models cannot represent either property. A response curve model
fits a nonlinear, parametric relationship between spend and KPI that
captures both.

------------------------------------------------------------------------

## A Common Mathematical Language: The Four-Parameter Family

All six response curves in `mrmopt` share the same four parameters,
giving them a unified interpretation regardless of curve type:

| Parameter | Name      | Interpretation                                         |
|----------------|----------------|-----------------------------------------|
| **b**     | Steepness | Controls how sharply the curve rises. Always negative. |
| **c**     | Floor     | Lower asymptote — the baseline KPI at zero spend.      |
| **d**     | Ceiling   | Upper asymptote — the maximum achievable KPI.          |
| **e**     | Midpoint  | The spend level at the inflection point of the curve.  |

The **range** of the curve — its total vertical travel — is

$$d - c$$

The floor `c` is rarely exactly zero in practice; it can absorb baseline
effects or organic response that occurs even without paid media.

The steepness parameter `b` governs the transition speed between floor
and ceiling. A very negative `b` produces a near step-function: nearly
all of the response is achieved within a narrow spend band. A `b` closer
to zero produces a slow, gradual rise.

------------------------------------------------------------------------

## Symmetry, Asymmetry, and the Six Curve Families

The six curves divide naturally into two groups based on where the
inflection point sits relative to the full curve range, and whether the
transition happens on a linear or logarithmic spend scale.

### Group 1: Linear-Scale Curves

These curves are defined in terms of raw spend

$$x$$

Their inflection point — the spend level where the curve changes from
accelerating to decelerating — is the parameter `e` directly.

#### Logistic (Symmetric)

$$y = c + \frac{d - c}{1 + e^{b(x - e)}}$$

The logistic curve is **symmetric** around its inflection point: the
curve rises at the same rate on either side of `e`. At

$$x = e$$

, the response is exactly at the midpoint of the range,

$$(c + d) / 2$$

#### Gompertz (Asymmetric — early inflection)

$$y = c + (d - c) \cdot e^{-e^{b(x - e)}}$$

The Gompertz curve is **asymmetric**: its inflection point occurs at a
lower response level than the midpoint of the range. Specifically, at

$$x = e$$

, the curve sits at

$$c + (d - c) / e \approx c + 0.368(d - c)$$

— roughly 37% of the way from floor to ceiling. This means the curve
rises quickly early and then flattens out gradually, spending more of
its range in the upper plateau than the lower ramp.

In media, this shape is often realistic: audiences are captured
relatively quickly at moderate spend levels, and the long tail of
incremental reach is expensive. The Gompertz's long upper tail models
this well.

#### Reflected Gompertz (Asymmetric — late inflection)

The reflected Gompertz mirrors the standard form, placing the inflection
point at approximately 63% of the range rather than 37%. It rises slowly
at first and accelerates before flattening — a shape that can represent
channels where significant investment is required before meaningful
response is generated (e.g., brand-building channels with long lags).

```{r fig-linear-curves, fig.cap = "The three linear-scale curves with identical parameters. The Gompertz inflects earlier (lower spend) than the logistic; the reflected Gompertz inflects later."}
x <- seq(0, 100000, length.out = 500)

curves_linear <- bind_rows(
  tibble(x = x, y = mrmopt:::rm_Logistic(x, b = -0.00012, c = 50, d = 900, e = 50000),
         type = "Logistic (symmetric)"),
  tibble(x = x, y = mrmopt:::rm_Gompertz(x, b = -0.00012, c = 50, d = 900, e = 50000),
         type = "Gompertz (early inflection)"),
  tibble(x = x, y = mrmopt:::rm_GompertzRef(x, b = -0.00012, c = 50, d = 900, e = 50000),
         type = "Reflected Gompertz (late inflection)")
)

ggplot(curves_linear, aes(x = x, y = y, colour = type)) +
  geom_line(linewidth = 1) +
  scale_x_continuous(labels = scales::dollar_format(scale = 1e-3, suffix = "k")) +
  labs(x = "Weekly Spend", y = "Conversions", colour = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Group 2: Log-Scale Curves

These curves replace raw spend

$$x$$

with

$$\log(x)$$

, so the inflection point and transition speed are expressed on a
multiplicative scale. The parameter `e` is now the spend level at the
inflection point (not a log-transformed value), but the shape of the
curve is governed by proportional changes in spend rather than absolute
ones.

**Practical consequence**: log-scale curves have monotonically
decreasing marginal returns from the first dollar — there is no initial
accelerating phase and no interior efficiency peak in the marginal
return curve. This makes them appropriate when diminishing returns set
in immediately, before the observed spend range.

#### Log-Logistic

$$y = c + \frac{d - c}{1 + e^{b(\log x - \log e)}}$$

The log-logistic is symmetric on the log scale: at

$$x = e$$

, response is at the midpoint

$$(c + d) / 2$$

. Doubling spend from `e/2` to `e` produces the same lift as doubling
from `e` to `2e`.

#### Weibull

$$y = c + (d - c) \cdot e^{-e^{b(\log x - \log e)}}$$

The Weibull is the log-scale analogue of the Gompertz: asymmetric, with
an early inflection. It models channels where significant response is
captured at relatively low spend levels on a proportional scale.

#### Reflected Weibull

The log-scale analogue of the reflected Gompertz — a slow early rise
with a late inflection on the log scale.

```{r fig-log-curves, fig.cap = "Log-scale curves evaluated over the same spend range. Whether marginal return is monotonically decreasing depends on the steepness parameter |b|; with these parameters (|b| = 2.5), all three have interior MR peaks."}
x_log <- seq(1000, 200000, length.out = 500)

curves_log <- bind_rows(
  tibble(x = x_log,
         y = mrmopt:::rm_LogLogistic(x_log, b = -2.5, c = 50, d = 900, e = 80000),
         type = "Log-Logistic (symmetric on log scale)"),
  tibble(x = x_log,
         y = mrmopt:::rm_Weibull(x_log, b = -2.5, c = 50, d = 900, e = 80000),
         type = "Weibull (early inflection on log scale)"),
  tibble(x = x_log,
         y = mrmopt:::rm_WeibullRef(x_log, b = -2.5, c = 50, d = 900, e = 80000),
         type = "Reflected Weibull (late inflection on log scale)")
)

ggplot(curves_log, aes(x = x, y = y, colour = type)) +
  geom_line(linewidth = 1) +
  scale_x_continuous(labels = scales::dollar_format(scale = 1e-3, suffix = "k")) +
  labs(x = "Weekly Spend", y = "Conversions", colour = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

------------------------------------------------------------------------

## Mathematical Properties Relevant to Media Planning

### Marginal Return

The **marginal return** (MR) is the derivative of the response curve
with respect to spend — the incremental KPI generated by the next dollar
of spend. It is the most operationally important quantity in media
planning, because it governs the efficiency of incremental investment.

For linear-scale curves (logistic, Gompertz, reflected Gompertz), the
marginal return curve itself has an interior maximum: there is a spend
level at which the marginal return is highest. Below that level, each
additional dollar is *more* efficient than the last (the accelerating
phase). Above it, each dollar is *less* efficient (the diminishing
returns phase). The spend at peak MR defines the lower bound of the
efficient investment range.

For log-scale curves, whether or not an interior MR peak exists depends
on the steepness parameter `b`. These curves are S-shaped on the log(x)
scale, but marginal return in dollars is:

$$\frac{dy}{dx} = \frac{dy}{d(\log x)} \cdot \frac{1}{x}$$

The first factor (the log-scale gradient) rises then falls, while the
second factor (1/x) decays monotonically. When **\|b\| \> 1**, the
log-scale gradient rises fast enough to overwhelm the 1/x decay,
producing an interior MR peak — qualitatively similar to linear-scale
curves. When **\|b\| ≤ 1**, the 1/x decay dominates and MR is
monotonically decreasing from the first dollar.

In practice, fitted log-scale curves often have \|b\| \> 1, so an
interior MR peak is common. When one exists, `mrmopt` uses the same
peak-based efficient range as linear-scale curves. When MR is genuinely
monotonically decreasing (no interior peak), `mrmopt` anchors range
bounds to the MR at current observed spend instead.

```{r fig-marginal-return, fig.cap = "Marginal return curves for a Gompertz and a Log-Logistic. Both have an interior MR peak with these parameters (|b| = 2.5 > 1 for the Log-Logistic)."}
delta <- 1

mr_gompertz <- diff(mrmopt:::rm_Gompertz(x, b = -0.00012, c = 50, d = 900, e = 50000)) / delta
mr_loglogistic <- diff(mrmopt:::rm_LogLogistic(x_log, b = -2.5, c = 50, d = 900, e = 80000)) / delta

mr_df <- bind_rows(
  tibble(x = x[-1],     mr = mr_gompertz,    type = "Gompertz"),
  tibble(x = x_log[-1], mr = mr_loglogistic, type = "Log-Logistic")
)

ggplot(mr_df, aes(x = x, y = mr, colour = type)) +
  geom_line(linewidth = 1) +
  scale_x_continuous(labels = scales::dollar_format(scale = 1e-3, suffix = "k")) +
  labs(x = "Weekly Spend", y = "Marginal Return (conversions / $)", colour = NULL) +
  facet_wrap(~type, scales = "free") +
  theme_minimal() +
  theme(legend.position = "none")
```

#### When does the MR peak disappear?

For the log-logistic, the MR peak location can be derived analytically.
At the peak, the sigmoid value satisfies $$p = (b - 1) / (2b)$$, which
only has a solution in (0, 1) when \|b\| \> 1. Below that threshold, the
1/x decay always dominates and MR decreases monotonically.

```{r fig-mr-vs-steepness, fig.cap = "Log-Logistic marginal return curves for different steepness values. When |b| > 1, there is an interior peak; when |b| ≤ 1, MR is monotonically decreasing."}
x_mr <- seq(1000, 200000, length.out = 2000)
b_demo <- c(-0.8, -1.0, -1.5, -2.5, -5.0)

mr_by_b <- purrr::map_dfr(b_demo, function(b_val) {
  y <- mrmopt:::rm_LogLogistic(x_mr, b = b_val, c = 50, d = 900, e = 80000)
  tibble(
    x  = x_mr[-1],
    mr = diff(y) / diff(x_mr),
    b  = paste0("b = ", b_val,
                if (abs(b_val) <= 1) "  (no peak)" else "  (has peak)")
  )
})

ggplot(mr_by_b, aes(x = x, y = mr, colour = b)) +
  geom_line(linewidth = 0.9) +
  scale_x_continuous(labels = scales::dollar_format(scale = 1e-3, suffix = "k")) +
  labs(x = "Weekly Spend", y = "Marginal Return (conversions / $)",
       colour = "Steepness") +
  theme_minimal() +
  theme(legend.position = "right")
```

### Asymptotes and the Ceiling

The ceiling parameter `d` is a **theoretical maximum** — the KPI the
channel would approach with unlimited spend. In practice, budgets rarely
get close to this ceiling, and posterior uncertainty around `d` tends to
be high unless data extends into the flat upper region of the curve.

This matters for prior specification: the ceiling prior should reflect
domain knowledge about how much response is plausibly achievable, not be
left entirely unconstrained. An unconstrained ceiling can allow the
posterior to place the inflection point far outside the data range,
producing a nearly linear fit that looks good in-sample but extrapolates
poorly.

### The Efficient Operating Range

Three spend levels partition the curve into interpretable zones:

- **Range min** (peak MR): The spend level where each dollar is most
  efficient. Spending below this level means leaving efficiency on the
  table.
- **Range max** (MR decay threshold): The spend level where marginal
  return has declined to a specified fraction of its peak (default 70%).
  Beyond this, diminishing returns accelerate meaningfully.

```{r fig-operating-range, fig.cap = "A Gompertz curve annotated with the three operating range points."}
x_range <- seq(0, 120000, length.out = 1000)
y_gompertz <- mrmopt:::rm_Gompertz(x_range, b = -0.00005, c = 50, d = 900, e = 50000)
mr_g <- c(NA, diff(y_gompertz))

peak_mr_x  <- x_range[which.max(mr_g)]
decay_mr   <- 0.7 * max(mr_g, na.rm = TRUE)
decay_x    <- x_range[which(mr_g <= decay_mr & x_range > peak_mr_x)[1]]

tibble(x = x_range, y = y_gompertz) |>
  ggplot(aes(x, y)) +
  geom_line(linewidth = 1, colour = "#2d6fa6") +
  geom_vline(xintercept = peak_mr_x, linetype = "dashed", colour = "#e07b39") +
  geom_vline(xintercept = decay_x,   linetype = "dashed", colour = "#c94040") +
  annotate("text", x = peak_mr_x, y = 150, label = "Peak MR\n(range min)",
           hjust = -0.1, size = 3, colour = "#e07b39") +
  annotate("text", x = decay_x,   y = 350, label = "70% MR decay\n(range max)",
           hjust = -0.1, size = 3, colour = "#c94040") +
  scale_x_continuous(labels = scales::dollar_format(scale = 1e-3, suffix = "k")) +
  labs(x = "Weekly Spend", y = "Conversions") +
  theme_minimal()
```

------------------------------------------------------------------------

## How Parameters Shape the Curve

Understanding how each parameter changes the curve shape helps with
prior specification and interpreting fitted models.

### Steepness (b)

`b` controls the transition speed. More negative values produce a
sharper inflection; values closer to zero produce a slower, more gradual
rise.

```{r fig-steepness, fig.cap = "Effect of steepness (b) on the Gompertz curve. All other parameters held constant."}
b_vals <- c(-0.00006, -0.00012, -0.00025, -0.0005)

map_dfr(b_vals, ~ tibble(
  x = x,
  y = mrmopt:::rm_Gompertz(x, b = .x, c = 50, d = 900, e = 50000),
  b = paste0("b = ", .x)
)) |>
  ggplot(aes(x, y, colour = b)) +
  geom_line(linewidth = 1) +
  scale_x_continuous(labels = scales::dollar_format(scale = 1e-3, suffix = "k")) +
  labs(x = "Weekly Spend", y = "Conversions", colour = "Steepness") +
  theme_minimal()
```

### Floor and Ceiling (c, d)

`c` shifts the entire curve up or down; `d` expands or contracts the
total range. Their difference $$d - c$$ is the total achievable lift
attributable to paid media.

```{r fig-ceiling, fig.cap = "Effect of ceiling (d) on the Gompertz curve. Floor held at c = 50."}
d_vals <- c(400, 700, 900, 1200)

map_dfr(d_vals, ~ tibble(
  x = x,
  y = mrmopt:::rm_Gompertz(x, b = -0.00012, c = 50, d = .x, e = 50000),
  d = paste0("d = ", .x)
)) |>
  ggplot(aes(x, y, colour = d)) +
  geom_line(linewidth = 1) +
  scale_x_continuous(labels = scales::dollar_format(scale = 1e-3, suffix = "k")) +
  labs(x = "Weekly Spend", y = "Conversions", colour = "Ceiling") +
  theme_minimal()
```

### Midpoint (e)

`e` slides the curve horizontally along the spend axis. A higher `e`
means the inflection point — and therefore the efficient operating range
— occurs at higher spend levels.

```{r fig-midpoint, fig.cap = "Effect of midpoint (e) on the Gompertz curve. The inflection point shifts right as e increases."}
e_vals <- c(20000, 40000, 60000, 80000)

map_dfr(e_vals, ~ tibble(
  x = x,
  y = mrmopt:::rm_Gompertz(x, b = -0.00012, c = 50, d = 900, e = .x),
  e = paste0("e = $", scales::comma(.x))
)) |>
  mutate(e = factor(e, levels = paste0("e = $", scales::comma(e_vals)))) |>
  ggplot(aes(x, y, colour = e)) +
  geom_line(linewidth = 1) +
  scale_x_continuous(labels = scales::dollar_format(scale = 1e-3, suffix = "k")) +
  labs(x = "Weekly Spend", y = "Conversions", colour = "Midpoint") +
  theme_minimal()
```

------------------------------------------------------------------------

## Choosing a Curve Type

No curve type is universally correct. The choice depends on the expected
shape of diminishing returns for the channel and what the data can
support.

| Curve | Inflection | Log scale? | When to consider |
|-----------------|-----------------|-----------------|----------------------|
| Gompertz | Early (\~37%) | No | Most digital channels; quick saturation after moderate spend |
| Logistic | Middle (50%) | No | Channels with symmetric response around a natural midpoint |
| Reflected Gompertz | Late (\~63%) | No | Brand-building or awareness channels with slow early response |
| Log-Logistic | Middle (log scale) | Yes | TV, OOH; proportional spend changes drive proportional response |
| Weibull | Early (log scale) | Yes | High-reach channels that saturate quickly on a proportional basis |
| Reflected Weibull | Late (log scale) | Yes | Channels requiring sustained investment before response builds |

**On model selection**: rather than selecting a single curve type a
priori, `mrmopt` supports fitting multiple types and comparing them
using Bayes R² and posterior predictive checks. See the *Model
Comparison & Diagnostics* vignette for guidance on this workflow.

::: callout-note
Log-scale curves require that spend is strictly positive in all observed
weeks. If your data contains zero-spend weeks, `mrmopt` can inject a
small offset automatically, but this should be reviewed — a channel with
true zero-spend weeks may be better modelled with a linear-scale curve.
:::

------------------------------------------------------------------------

## Data Requirements

A response curve model is only as good as the data used to fit it. There
are three fundamental requirements.

### 1. Spend Variation

The most important requirement. The model learns the shape of the
saturation curve by observing how response changes as spend changes. If
spend is nearly constant across all observed weeks, the data cannot
identify the curve — all parameter combinations that pass through that
narrow spend band are equally consistent with the data.

A practical minimum is a spend range spanning **at least one-half to one
order of magnitude** (e.g., lowest week no more than 50% of the highest
week). Greater variation provides more information and reduces posterior
uncertainty.

```{r fig-spend-variation, fig.cap = "Spend variation in the mrmopt_data example dataset. Adequate variation is present across all channels.", fig.height = 5}
data(mrmopt_data)

ggplot(mrmopt_data, aes(x = week, y = spend)) +
  geom_line(colour = "#2d6fa6", linewidth = 0.6) +
  facet_wrap(~channel, scales = "free_y", ncol = 2) +
  scale_y_continuous(labels = scales::dollar_format(scale = 1e-3, suffix = "k")) +
  labs(x = NULL, y = "Weekly Spend") +
  theme_minimal()
```

### 2. Series Length

Fitting a Bayesian nonlinear model requires enough observations to
constrain four parameters. As a rough guide:

- **Minimum**: \~52 weeks (one year). Below this, posteriors will be
  wide and sensitive to prior choice.
- **Recommended**: 104 weeks (two years) or more.
- **Longer is better** — but only if the underlying relationship is
  stable across the entire period. Two years of data from a channel that
  changed its creative strategy or targeting mid-period may be worse
  than one year of stable data.

### 3. Structural Stability

The model assumes a single, stable response curve across the entire
observed period. Structural breaks — a major change in audience
targeting, a shift in channel strategy, a platform algorithm change —
violate this assumption. Fitting through a break produces a curve that
represents neither period well.

`mrmopt` provides two approaches for handling within-channel structural
variation:

-   **Hierarchical models** via `fit_response_hier()`: When a channel
    has a natural grouping structure (e.g., creative variants, audience
    segments, or partner stations), a hierarchical model fits a shared
    curve shape with group-level deviations. This borrows strength
    across groups while allowing each to have its own effective
    saturation behaviour. See the *Hierarchical Curves* vignette for
    details.

-   **Rolling-window models** via `mrm_rolling()` (planned):
    For channels where the response curve itself drifts over time — due
    to seasonality, competitive dynamics, or platform changes — a
    rolling-window approach fits separate curves to overlapping time
    windows, tracking how parameters evolve. This is on the development
    roadmap. For *pooled* per-period estimates available today, use
    `fit_response_hier()` with a time bucket as the `group`.

If breaks are suspected and neither approach applies, fit separate
models to each sub-period. The `date_range` metadata stored on each
fitted model (accessible as `fit$date_range`) makes it straightforward
to compare models across periods using `mrms_plot_compare()`.

------------------------------------------------------------------------

## Summary

The six response curves in `mrmopt` share a unified four-parameter
structure. The key distinctions are:

- **Symmetry**: Logistic and Log-Logistic are symmetric around their
  inflection point; the Gompertz family is asymmetric.
- **Scale**: Linear-scale curves (Logistic, Gompertz, Reflected
  Gompertz) always have an interior marginal return peak. Log-scale
  curves (Log-Logistic, Weibull, Reflected Weibull) have an interior MR
  peak when \|b\| \> 1 and monotonically decreasing MR when \|b\| ≤ 1.
- **Inflection timing**: Within each scale group, the standard,
  Gompertz, and reflected Gompertz variants place the inflection at 50%,
  \~37%, and \~63% of the response range respectively.

Fitting begins with `fit_response()`. See
[Getting Started](getting_started.html) for the full workflow, and
[Priors & Model Tuning](priors_and_tuning.html) for guidance on how these
mathematical properties inform prior specification.
