---
title: "Time-Varying Response Curves"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Time-Varying Response Curves}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 8,
  fig.height = 4.5,
  message  = FALSE,
  warning  = FALSE
)
library(mrmopt)
```

Standard response curve fitting treats the whole observation window as a
single stationary period. In practice, the relationship between spend and KPI
drifts — audiences change, creative fatigues or improves, competitive pressure
shifts. `fit_response_tv()` models one or more of the curve parameters
(`b`, `c`, `d`, `e`) as smooth or Gaussian-process functions of time instead
of constants, so the curve itself can evolve across the fitted period.

This vignette walks through the API on a small simulated series — a handful
of chains and a short chain length, purely to keep the examples fast. It is
not a statistical validation; see `dev/tv_smooth_curves.R` and
`dev/tv_smooth_params.R` for the real-data evidence behind this feature's
defaults.

---

## The data

A single channel, 60 weekly observations, where the midpoint (`e`) — the
spend level at which the response curve is half-saturated — drifts down over
the year: the same conversion rate is reached at less spend by the end of the
period than at the start (e.g. an improving creative, or a maturing
audience). The ceiling, floor, and steepness are held fixed so the drift in
`e` is the only thing moving.

```{r simulate}
set.seed(42)
n <- 60
t <- seq(0, 1, length.out = n)

true_b <- -0.004
true_c <- 10
true_d <- 200
true_e <- 800 - 400 * t   # drifts from 800 down to 400 over the year

spend <- runif(n, 100, 1500)
gompertz <- function(x, b, c, d, e) c + (d - c) * exp(-exp(b * (x - e)))
kpi <- pmax(gompertz(spend, true_b, true_c, true_d, true_e) + rnorm(n, 0, 8), 0)

channel_data <- data.frame(
  spend = spend,
  kpi   = kpi,
  date  = as.Date("2023-01-02") + (seq_len(n) - 1) * 7
)
head(channel_data)
```

---

## When to reach for this instead of `fit_response_hier()`

Partial pooling of curve parameters across time buckets already works with
no new function — `fit_response_hier(group = "quarter", pool = "d")` treats a
time bucket exactly like a sub-channel. It is a reasonable first thing to
try, and needs no smooth/GP machinery. `fit_response_tv()` is the better
choice when:

- bucket boundaries would be arbitrary and you want a continuous trajectory
  instead of a step function, or
- there are too few time buckets to safely pool more than one parameter —
  hierarchical pooling of shape parameters can fail outright when few
  buckets remain to estimate multiple group-level variances (see
  `ROADMAP.md`'s Known Limitations for `fit_response_hier()`'s time-bucket
  use), or
- the parameter you care about (the midpoint, here) is the one most likely
  to benefit — see the next section.

---

## Fit the model

```{r fit-main}
fit <- fit_response_tv(
  channel_data,
  spend = "spend", kpi = "kpi", date = "date",
  varying = "e",        # the default; see "Choosing varying" below
  method  = "spline",   # the default; s(t)
  chains  = 2, iter = 1500, warmup = 750, refresh = 0
)
```

`varying = "e"` is the package default, not an arbitrary starting point —
across the channels tested during development, letting the midpoint vary was
the only choice that converged cleanly on all of them; letting steepness
(`b`) vary, or letting two parameters vary jointly, was fine on wide-spend
channels but risked real convergence trouble on a narrow one — see "A
flagged fit" below.

```{r print-fit}
print(fit)
```

`print()` surfaces two checks, not just the parameter trajectory:

- a **pre-fit** spend-variability heuristic (advisory — built from limited
  evidence, not a certainty)
- a **post-fit** convergence check (R-hat, effective sample size,
  divergences, treedepth) — this one is authoritative. A clean pre-fit ratio
  does not guarantee a clean post-fit result, and vice versa; both are
  always shown.

Both are also available directly:

```{r diagnostics}
fit$identifiability
fit$diagnostics
```

---

## Visualization

`mrm_plot_tv()` has two views. `type = "trajectory"` plots the varying
parameter(s) against time:

```{r plot-trajectory}
mrm_plot_tv(fit, type = "trajectory")
```

`type = "evolution"` shows the fitted curve itself changing shape — a
continuous heatmap (time on the x-axis, spend on the y-axis, reading
left-to-right like an ordinary time series) alongside a few discrete curve
snapshots colored by date:

```{r plot-evolution, fig.height=5}
mrm_plot_tv(fit, type = "evolution")
```

The recovered trajectory should track the simulated drift in `true_e` — the
curve visibly saturates at lower spend by the end of the series than at the
start.

---

## A flagged fit

Not every `varying`/data combination is safe, and the point of the
identifiability gates is to say so rather than silently return a number.
Here the same shape of series, but with spend confined to a narrow band and
`b` (steepness) instead of `e` set to vary — the two conditions found,
during development, to risk convergence trouble:

```{r fit-flagged}
narrow_data <- channel_data
narrow_data$spend <- runif(n, 900, 1100)   # spend barely varies

fit_flagged <- fit_response_tv(
  narrow_data,
  spend = "spend", kpi = "kpi", date = "date",
  varying = "b",
  chains = 2, iter = 1500, warmup = 750, refresh = 0
)
print(fit_flagged)
```

The **pre-fit** heuristic flags this one (spend ratio well under the
stricter threshold `varying = "b"` triggers). The **post-fit** gate happens
to come back clean here — but look at the Bayes R2: it collapses to near
zero. With spend barely varying, there is no information to estimate how
steepness moves over time, so the chains agree there's nothing to find
rather than disagreeing about where the answer is; R-hat alone would not
have caught this. That is exactly why both checks exist and neither is
sufficient alone — a low R2 on a "clean" fit is as much a red flag as a bad
R-hat. On real data, this same failure mode (narrow spend range, steepness
or multiple parameters varying) has produced genuine posterior
multimodality — chains landing in different, disagreeing answers rather
than agreeing there's no answer — which the post-fit gate does catch; see
`dev/tv_paid_search_fix.R` for that case. Either way, the fix is the same:
a smaller `varying` set, `method = "gp_approx"`, or falling back to
`fit_response()` if the channel's own spend range cannot support time
variation at all.

---

## Static curves for optimization

A `mrmfit_tv` doesn't itself feed `opt_mix()` — its curve is a function of
time, not a single point. `mrm_tv_snapshot()` extracts a static curve at a
specific date, with full posterior uncertainty carried through (not just the
point estimate), so it behaves like an ordinary `fit_response()` fit for
every purpose `opt_mix()` needs:

```{r snapshot}
snap <- mrm_tv_snapshot(fit, at = "latest")
print(snap)
```

```{r opt-mix}
opt_mix(list(channel = snap), budget = mean(channel_data$spend) * 52, verbose = FALSE)
```

Pass `date = ` instead of `at = ` to snapshot at a specific point in the
trajectory rather than the most recent one.

---

## Choosing `varying` and `method`

| `varying` | Notes |
|---|---|
| `"e"` (default) | Midpoint. The safest single choice found during development — converged cleanly across every tested channel. |
| `"d"` | Ceiling. Close behind `"e"` in reliability; a reasonable alternative when the ceiling, not the saturation point, is what's expected to move. |
| `"b"` | Steepness. Weaker fit and the parameter most prone to multimodality on narrow-spend channels — see above. |
| more than one | Not unsafe by default, but not obviously better either: on the channels tested, letting two parameters vary jointly rarely beat the best single-parameter choice by enough to justify the extra identifiability risk. |

| `method` | Cost (measured on one real channel) | When to use |
|---|---|---|
| `"spline"` (default) | ~76s, 0 divergences | Default. `s(t)`. |
| `"gp_approx"` | ~77s, 0 divergences | A same-cost alternative to spline; Hilbert-space approximate GP. |
| `"gp"` | ~2,914s (49 min), 25% max-treedepth | Exact GP. Available, not recommended as a default — meaningfully slower *and* a real sampler-geometry problem, not just slower. |

---

## Where to go next

| Topic | Vignette |
|-------|----------|
| Partial pooling of curve parameters, including across time buckets | [Hierarchical Models](hierarchical_models.html) |
| Budget optimization objectives and constraints | [Optimization](optimization.html) |
| Mathematical foundations of the six curve forms | [Response Curve Theory](response_curve_theory.html) |
